Analysing the three-dimensional structure of {\em P. falciparum}

### 2017-07-26 doc/manuscript/paper.tex
More on the paper

%% Use natbib with these recommended options
\usepackage[authoryear,comma,longnamesfirst,sectionbib]{natbib}

@@

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effects of malaria extend far beyond direct
measures of mortality and the disease is thought to cost more than US\$21
billion a year \citep{onwejkwe:do}.

Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
its host through a mosquito bite. Out of the 5 species that can infect human,
{\em P. falciparum} is by the far the most deadly with 99\% of the deaths
associated to it. The parasite has a complex life cycle, with multiple stages
both in the human and mosquito host. The human host is infected with
\textit{sporozoites} through the bite of a infected female  {\em Anopheles}
mosquito. The \textit{sporozoites} quickly migrate to the liver, where they
start a two weeks multiplication process. They are then released as
\textit{merozoites} in the bloodstream and proceed to infect red blood cells. The
parasites then start their ``erythrocytic'' cycle (through \textit{rings},
\textit{trophozoites} and \textit{schizonts} stages), via another round of
replications in red blood cells, through an unusual process of cell division
called \textit{schizogony}. The parasite first undergoes multiple rounds of nuclear
replication and division into 13 to 32 daughter cells, until the red blood
cell bursts and provokes the release of \textit{merezoites} and the infective cycle
starts anew. This asexual replication cycle is responsible for the symptoms
and the complications of the disease: anemia, tertian fever, \dots

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citet{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms . Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at the
epigenetic and post-translational levels
\citep{cui:chromatin-mediated, duffy:role, hoeijmakers:placing,
horrocks:control, deitsch:mechanisms}.

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus. Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes. The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citet{duan:three-dimensional,
burton:xx}, bacteria, flies, plants and numerous human and mouse cell lines.
Within the past few years, two studies of the three-dimensional structure of
{\em P. falciparum} have been released.  \citet{lemieux:genome-wide} probed
several strains of {\em P. falciparum} to understand the link between 3D
structure and the complex regulation of \textit{var} genes, a family of genes
involved the invasion of red blood cell and also responsible for the
parasite's great capacity to evade immuno-therapy.
\citet{ay:three-dimensional} assayed the 3D structure of the genome of the
parasite during three key stages of the erythrocytic cycle.
\citet{ay:multiple} reviews the key biological findings of those studies.

@@

\todo{A revoir}

In this section, we review ``data-driven'' methods to infer 3D models of DNA
from Hi-C data. These methods broadly fall into two categories: (i)
\textit{consensus} approaches, that aim at inferring a unique mean structures
best representing the contact count data; (ii) \textit{ensemble} methods that
yield a population of structures.

@@

After introducing some notations and common terminology, I will dive both in
consensus and population inference methods.

\subsection*{Notation}

A typical 3C/Hi-C experiment can be summarized as a contact
count matrix $\mathbf{C} \in \RR^{n \times n}$, where each row and column
correspond to a genomic-window (or loci), and each entry $c_{ij}$ to the
number of times loci $i$ and $j$ have been observed interacting. Contact count
matrices are subjected to biological and technical biases that need to be
corrected \citep{imakaev:iterative, cournac:normalization, yaffe:probabilistic,
hu:hicnorm}.

Methods model chromosomes as a series of beads in 3D, each bead corresponding
to a specific genomic window. We denote by $\textbf{X} \in \RR^{3 \times n}$
the coordinate matrix of the structure(s), $x_i \in \RR^3$ corresponds to
the coordinate matrix of the i-th beads and $d_{ij} = \|x_i - x_j\|_2$ the
euclidean distance separating beads $i$ and $j$.

@@

Consensus methods aim at inferring a model that best represents the contact
count matrix. They model each chromosome by a chain of beads, and attempt to
place the beads in a 3D space to minimize an objective function
$\mathcal{O}(\mathbf{O}, \mathbf{C})$, sometimes
under constraints.

\begin{equation*}
\renewcommand{\arraystretch}{2}
\begin{array}{ccl}
\underset{\Xb}{\text{minimize}} & & \mathcal{O}(\mathbf{C}, \Xb)
\end{array}
\end{equation*}

@@

Early methods \citep{dekker:capturing, duan:three-dimensional,
tanizawa:mapping, ay:three-dimensional} cast the 3D inference problem as a
multidimensional scaling (MDS) problem: chromosomes are models as a chain of
beads, to place in 3D such that the distance between bead $i$ and $j$ matches
as closely as possible a ``wish-distance'' $\delta_{ij}$ derived from contact
counts.

@@

The first step is thus to find an adequate count-to-distance mapping to
convert contact counts into pairwise wish-distances $\delta_{ij}$.
\citet{dekker:capturing} models the 78 pairwise contact count of {\em S.
cerevisiae}'s chromosome III as a Worm-like chain polymer,
\citet{duan:three-dimensional} uses a linear mapping between contact counts
and wish distances, \citet{ay:three-dimensional} relies on the biophysical
properties of fractal globule polymers, \citet{tanizawa:mapping} inferred the
count-to-distance mapping by fitting the relationship using known pairwise
distances obtained through high-resolution FISH measures. While parameters of
these counts-to-distance functions vary, most of them can be summarized as a
power-law relationship: $c_{ij} = \beta d_{ij}^\alpha$.

@@

A crucial step of MDS-based method is the conversion of counts into
wish-distances. As described earlier, converting contact counts into
wish-distances require strong assumptions that may not be met in practice. For
example, this mapping changes from one organism to another
\citep{fudenberg:higher-order}, from one resolution to another
\citep{zhang:inference} or even from one time point to another during the cell
cycle \citep{le:high-resolution, ay:three-dimensional}. Several methods have
since been proposed to alleviate this problem.

\citet{ben-elazar:spatial} casts the inference as a \emph{non-metric MDS}
\citep{kruskal:multidimensional}, where the 3D structure is inferred jointly
with the wish-distances. The authors first filter the
interaction counts to keep only the most significant interactions and
interpolate the missing values to obtain a smooth, symmetric positive definite
matrix.

\citet{zhang:inference} proposes to parametrize the count-to-distance mapping
as a power-law ($d = \beta c^\alpha)$, and to infer jointly the parameters
jointly with the structure. In addition, the authors propose to add a
penalization term for non-interacting beads. The proposed penalization allows
to cast the problem as a convex optimization and thus guarantees to retrieve
the optimal structure.

@@

@@

\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
the counts-to-distance transfer function and infer a genome structure that
best explains the observed data.

@@

{\em Ensemble approaches} aim at inferring a population of structures
representative of the contact count map. The methods fall into two distinct
categories: the first type of problem casts a non-convex or ill-defined
optimization problem and sample local minima of the function
\citep{bau:three-dimensional, umbarger:three-dimensional}, while the second
type proposes a statistical modeling of the problem and samples the posterior
distribution \citep{rousseau:three, hu:bayesian}.A parallel can be made
between the first group of methods and {\em Consensus} MDS type of approaches,
and the second with {\em Consensus} statistical based methods.

@@

Similarly to \citet{varoquaux:statistical}, \citet{rousseau:three} and
\citet{hu:bayesian} proposes to model contact counts with a formal
probabilistic model. \citet{rousseau:three} models observed contact counts
$c_{ij}$ as a random Gaussian variable of mean $\beta d_{ij}^{\alpha}$,
$\alpha \leq 0$ and variance $\sigma_{ij}$ estimated directly from the contact
count data, while \citet{hu:bayesian} models contact counts as random Poisson
variables of mean $\beta d_{ij}^\alpha$. The authors then sample from the
posterior using MCMC.

@@

\begin{table}[ht!]
\begin{center}
\footnotesize
\begin{tabular}{rlccccc}
\hline
\multirow{2}{*}{\textbf{\footnotesize Publication}} & \multirow{2}{*}{\textbf{\footnotesize Name}} &
\textbf{\footnotesize Consensus or} 
&\multirow{2}{*}{\textbf{\footnotesize MDS-based}} & \textbf{\footnotesize Statistical} &
\multirow{2}{*}{\textbf{\footnotesize Available}} \\
& & \textbf{\footnotesize Ensemble}
 & & \textbf{\footnotesize model} & \\
\hline
\hline
\footnotesize{\cite{dekker:capturing}} & & C & & & \\
\footnotesize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
\footnotesize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
\footnotesize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
\footnotesize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
\footnotesize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
\footnotesize{\cite{bau:three-dimensional}} & & E & & &  \\
\footnotesize{\cite{umbarger:three-dimensional}} & & E  & & &\\
\footnotesize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
\footnotesize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
\footnotesize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
\footnotesize{\cite{kalhor:genome}} & & E &   & &\\
\footnotesize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
\footnotesize{\cite{trieu:large}} & & C & & & \\
\footnotesize{\cite{trieu:3D}} & & C & &  & \\
\footnotesize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
\footnotesize{\cite{nagano:single-cell}} & & E & & & \\
\footnotesize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
\hline
\end{tabular}
\end{center}
\caption{\bf A comparison of 3D inference methods}{In this table, we summarize
properties of published methods to infer the 3D structure of the genome: (1)
is it a consensus or a ensemble based inference? (2) Is it an MDS based
method? (3) or relies on a statistical modeling; (4) is the software
available or not (to the best of our knowledge).}
\end{table}

\section*{Modeling DNA as a flexible fibers under constraints}

@@

The budding yeast \textit{S. cerivisiae}'s 3D structure has been extensively
studied, both through 3C-type studies \citep{dekker:capturing,
duan:three-dimensional, burton:} and through bio-imaging experiments
\citep{XX}. The small size of its genome, the well-known hallmarks of its
genome architecture and the availability of high resolution contact maps and
FISH data set quickly led several teams to investigate the minimal set of
constraints needed to reproduce the hallmarks of its genome architecture.

\cite{tjong:physical, tokuda:dynamical, tjong:physical} model {\em S.
cerevisiae}'s chromosomes as flexible random fibers under a small set of
constraints. While the exact modeling proposed by the three groups differ, the
set of constraints are similar: \textit{(i)} the chromosomes are constrained
into a spherical ball representing the nucleus; \textit{(ii)} centromeres are
constrained into a spherical ball tethered to the nuclear membrane;
\textit{(iii)} telomeres are tethered to the nuclear membraine; \textit{(iv)}
rDNA is constrained into the nucleolus, represented as a spherical ball
opposite to the centromeres.

@@

In the end, the objective function minimized is the following:

\begin{equation}
\end{equation}

@@

\subsection*{\textit{P. falciparum}'s 3D structure cannot be explained by a
constrained volume exclusion model}

Running a similar model on {\em P. falciparum}

@@

\subsection*{Profiling {\em P. falciparum}'s genome architecture during the
asexual cycle}
\label{sec:data}

In the past year, two groups have published novel dataset studying {\em P.
falciparum}'s 3D genome architecture using Hi-C. \citet{lemieux:genome-wide}
studied several strains associated with populations expression unique
\textit{var} genes, in order to study key folding properties of the 3D
structure relating to \textit{var} gene expression. On the other hand,
\citet{ay:three-dimensional} focus three key time-point of the development
cycle. Table~\ref{tab:data} summarizes the available data sets and their key
properties after processing the data as described in
\citet{ay:three-dimensional} and combining all available libraries for each
life-cycle stage and strain.

\subsection*{Structure stability}

Both the constrained and unconstained problem stated above are not convex. The
reader may well ask how sensitive the result is to initialize. In short, it is
not. To study the stability of the results with respect to the initialization,
I perform for each stage presented here 1000 optimization, starting from
different initialization.

The first question that arise is simply: are structures from the same time
points but from a different start ponit more alike than structures from
different time points? To answer this question, I compute for each structure
and each time point a feature matrix, composed of a subsampled pairwise
euclidean distances of each structure (so that the resulting feature matrix
fits in memory on a laptop). I then compute a stochastic PCA on this feature
matrix and plot the projection in 2D space using the 3 first components. The
results clearly show that structures from the same time point cluster
strongly. Repeating this experiment on singly chromosomes yields the same
result.

The second question one can ask is how do the structures differ. Tackling this
question is very challenging, but can be reformulated in a much easier way.
Are the hallmarks of interest conserved across structures of a same stage?
\citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both identified
{\em P. falciparum} folded in very specific ways, with VRSSM genes highly
interacting. \citet{ay:three-dimensional} also observed strong clustering of
the centromeres, and enrichment in interaction at the telomeres. These
observation can lead to a rigourous approach at identifying whether families
of loci clustered in the structures and even quantifying this in a rigourous
manner.

@@

To assess whether groups of genes are colocolized in a 3D model,
\citet{ay:three-dimensional} leveraged a statistical method developped by
\citet{witten:assessment}, which requires labeling each pair of loci in two groups:
"close" or "far".

\fixme{This is unclear. Introduce notatinos and rewrite this section}

The authors used varying distance
thresholds (10\%, 20\% and 40\% of the nuclear diameter) to deem a locus pair
“close” and labeled all remaining
pairs in the set as “far”.  The authors then compare the enrichment of loci
pairs of a group being close and far by resampling loci amongst a same
chromosome.

@@

Applying MEPD to the Ring stage, \citet{capurso:distance-based} confirm that
centromers, telomers, VRSM genes (both overall, subtelomeric and internal)
colocalize.

\begin{table}
\scriptsize
\centering
\begin{tabular}{cccccccc}
\hline
\textbf{Name} & \textbf{Strain} & \textbf{Stage} & \textbf{Resolution} &
\textbf{Number of contacts} & \textbf{Perc. of \textit{cis}} & \textbf{Perc of
\textit{trans}}& \textbf{Reference} \\
\hline
\hline
Ay-\textit{rings} & 3D7 & Late Rings & 10~kb & 16711552 & 43\% & 57\% &  \cite{ay:three-dimensional} \\
Ay-\textit{trophozoites} & 3D7 & Trophozoites &10~kb & 56348498 & 53\% & 47\% & \cite{ay:three-dimensional} \\
Ay-\textit{schizonts} & 3D7 & Schizonts & 10~kb & 11652832 & 55\% & 45 \% & \cite{ay:three-dimensional} \\
Lemieux-\textit{A4+} & IT/BC6+ & Rings & 25~kb & 18488252 & 19\% & 81\% & \cite{lemieux:genome-wide} \\
Lemieux-\textit{A4} & IT/3G8 & Rings &  25~kb & 19674672 & 28\% & 72\% & \cite{lemieux:genome-wide}\\
Lemieux-\textit{A44} & IT/BC6- & Rings & 25~kb & 18660594 & 25\% & 75\% & \cite{lemieux:genome-wide}\\
Lemieux-\textit{DCJ\_On} & NF54/DCJ on & Rings & 25~kb & 3098370 & 26\% & 74\% &\cite{lemieux:genome-wide} \\
Lemieux-\textit{DCJ\_Off} & NF54/DCF Off & Rings & 25~kb & 2533470 & 26\% & 73\% &\cite{lemieux:genome-wide} \\
Lemieux-\textit{B15C2} & NF54/B15C2 & Rings & 25~kb &  1022996 & 12\% & 88\% & \cite{lemieux:genome-wide}\\
\hline
\end{tabular}
\caption{Summary of the available {\em P. falciparum} Hi-C datasets}
\end{table}

@@

>>>>

%% Use natbib with these recommended options
%\usepackage[authoryear,comma,longnamesfirst,sectionbib]{natbib} 
\usepackage{natbib}

@@

With more than 217 millions cases nearly 429 000 deaths in 2015,  malaria
remains a major disease burden in tropical and subtropical countries and an
impediment to economic development. In Africa, where more than 90\% of the
cases and deaths occurs, the effects of malaria extend far beyond direct
measures of mortality and the disease is thought to cost more than US\$21
billion a year \citep{onwujekwe:do}.
Malaria is caused by small unicellular protozoan {\em Plasmodium}, infecting
its host through a mosquito bite. Out of the 5 species that can infect human,
{\em P. falciparum} is by the far the most deadly with 99\% of the deaths
associated to it. The parasite has a complex life cycle, with multiple stages
both in the human and mosquito host. The human host is infected with
\textit{sporozoites} through the bite of a infected female  {\em Anopheles}
mosquito. The \textit{sporozoites} quickly migrate to the liver, where they
start a two weeks multiplication process. They are then released as
\textit{merozoites} in the bloodstream and proceed to infect red blood cells. The
parasites then start their ``erythrocytic'' cycle (through \textit{rings},
\textit{trophozoites} and \textit{schizonts} stages), via another round of
replications in red blood cells, through an unusual process of cell division
called \textit{schizogony}. The parasite first undergoes multiple rounds of nuclear
replication and division into 13 to 32 daughter cells, until the red blood
cell bursts and provokes the release of \textit{merezoites} and the infective cycle
starts anew. This asexual replication cycle is responsible for the symptoms
and the complications of the disease: anemia, tertian fever, \dots

While efficacious vaccines still remains a hope and drug resistance to anti
malaria continues to arise, genomic-based research on malaria seeks new
avenues for developing novel therapies \citep{kirchner:recent}. One of the
limiting factors of developments of new drugs is our poor understanding of the
mechanisms underlying the parasite's complex life cycle. While the development
of {\em P. falciparum} through the different stages of its life is thought to
be driven by coordinated changes in gene expression, the relative paucity of
transcription factors points to unusual gene regulatory mechanisms. Meanwhile,
the relative abundance of proteins related to chromatin structures, mRNA decay
and translation rates suggests alternative mechanisms of gene regulation at the
epigenetic and post-translational levels
\citep{cui:chromatin-mediated, duffy:role, hoeijmakers:placing,
horrocks:control, deitsch:mechanisms}.

In recent year, chromosome conformation captures-like method, broadly referred
to as Hi-C allowed to identify physical interactions between two regions in a
genome-wide fashion, yielding information on their relative spatial distance
in the nucleus. Hi-C has opened new avenues for more systematic analysis of
the three-dimensional folding of the genome, paving the way for a better
understanding of the relations between 3D structure and gene regulation,
replication timing, epigenetic changes as well as many other biological
processes. The 3D genome of a wide variety of organisms has been studied in
the past year, including many species of yeasts \citep{duan:three-dimensional,
burton:species-level, mizuguchi:cohesin-dependant}, bacteria
\citep{umbarger:three-dimensional}, flies \citep{sexton:three-dimensional},
plants \citep{feng:genome-wide, wang:genome-wide} and numerous human and mouse
cell lines \citep{lieberman-aiden:comprehensive, rao:3D}.

Within the past few years, two studies of the three-dimensional structure of
{\em P. falciparum} have been released.  \citet{lemieux:genome-wide} probed
several strains of {\em P. falciparum} to understand the link between 3D
structure and the complex regulation of \textit{var} genes, a family of genes
involved the invasion of red blood cell and also responsible for the
parasite's great capacity to evade our immune system.
\citet{ay:three-dimensional} assayed the 3D structure of the genome of the
parasite during three key stages of the erythrocytic cycle.
\citet{ay:multiple} reviews the key biological findings of those studies.

@@

The development of genowe-wide and high-throughput protocols to probe samples
for their 3D genome architecture naturally paved the way to systematic and in
depth study of the folding mechanisms of DNA.

Over the past 10 years, the challenge of building 3D models from contact maps
have been tackled by a plethora of methods. These model chromosomes as a
series of beads, and attempt to place the beads in a 3D euclidean space to
well represent the contact map. ``data-driven'' methods broadly fall into two
categories: (i) \textit{consensus} approaches, that aim at inferring a unique
mean structures best representing the contact count data; (ii)
\textit{ensemble} methods that yield a population of structures.

@@

@@

Consensus methods aim at inferring a model $\mathbf{X} \in \RR^{3 \times n}$
that best represents the contact count matrix $\mathbf{X} \in \RR^{3 \times
n}$, usually through the minimization of an objective function
$\mathcal{O}(\mathbf{O}, \mathbf{C})$, sometimes under constraints.

\begin{equation*}
\renewcommand{\arraystretch}{2}
\begin{array}{ccl}
\underset{\Xb}{\text{minimize}} & & \mathcal{O}(\mathbf{C}, \Xb) \,,
\end{array}
\end{equation*}

@@

Early methods cast the 3D inference problem as a
multidimensional scaling (MDS) problem: chromosomes are models as a chain of
beads, to place in 3D such that the distance between bead $i$ and $j$ matches
as closely as possible a ``wish-distance'' $\delta_{ij}$ derived from contact
counts \citep{dekker:capturing, duan:three-dimensional,
tanizawa:mapping, ay:three-dimensional} .

@@

The first step is thus to find an adequate count-to-distance mapping to
convert contact counts into pairwise wish-distances $\delta_{ij}$. 
\citet{dekker:capturing} models the 78 pairwise contact count of {\em S.
cerevisiae}'s chromosome III as a Worm-like chain polymer,
\citet{duan:three-dimensional} uses a linear mapping between contact counts
and wish distances, \citet{ay:three-dimensional} relies on the biophysical
properties of fractal globule polymers, \citet{tanizawa:mapping} inferred the
count-to-distance mapping by fitting the relationship using known pairwise
distances obtained through high-resolution FISH measures.

@@

A crucial step of MDS-based method is the conversion of counts into
wish-distances. As described earlier, converting contact counts into
wish-distances require strong assumptions that may not be met in practice. For
example, this mapping changes from one organism to another
\citep{fudenberg:higher-order}, from one resolution to another
\citep{zhang:inference} or even from one time point to another during the cell
cycle \citep{le:high-resolution, ay:three-dimensional}. To alleviate this
problem, after filtering interaction counts based on significance and
interpolation of missing values, \citet{ben-elazar:spatial} casts the
inference as a \emph{non-metric MDS} \citep{kruskal:multidimensional}, where
the 3D structure is inferred jointly with the wish-distances. Another idea is
to parametrize the count-to-distance mapping as a power-law ($d = \beta
c^\alpha)$, and to infer jointly the parameters jointly with the structure
\citet{zhang:inference}.

@@

\fixme{maybe skip this section}

@@

\texttt{Pastis} can automatically adjust the parameters $\beta$, $\alpha$ of
the counts-to-distance transfer function and infer a genome structure that
best explains the observed data. The strength of this methods comes from the
robustness to low signal-to-noise ratio.

\fixme{add results on the P. falciparum to show pastis's results on 
super noisy data}

@@

{\em Ensemble approaches} aim at inferring a population of structures
representative of the contact count map. The methods fall into two distinct
categories: the first type of problem casts a restraint-based 
optimization problem and sample local minima of the function
\citep{bau:three-dimensional, umbarger:three-dimensional}, while the second
type proposes a statistical modeling of the problem and samples the posterior
distribution \citep{rousseau:three, hu:bayesian}. A parallel can be made
between the first group of methods and {\em Consensus} MDS type of approaches,
and the second with {\em Consensus} statistical based methods.

@@

Similarly to \citet{varoquaux:statistical}, \citet{rousseau:three} and
\citet{hu:bayesian} proposes to model contact counts with a formal
probabilistic model. \citet{rousseau:three} models observed contact counts
$c_{ij}$ as a random Gaussian variable of mean $\beta d_{ij}^{\alpha}$,
$\alpha \leq 0$ and variance $\sigma_{ij}$ estimated directly from the contact
count data, while \citet{hu:bayesian} models contact counts as random Poisson
variables of mean $\beta d_{ij}^\alpha$. The authors then sample from the
posterior using MCMC. Obtaining a consensus structure from such a method can
be accomplished by selecting the maximum a posteriori.

@@

\begin{table}[ht!]
\begin{center}
\scriptsize
\begin{tabular}{rlccccc}
\hline
\multirow{2}{*}{\textbf{\footnotesize Publication}} & \multirow{2}{*}{\textbf{\footnotesize Name}} &
\textbf{\footnotesize Consensus or} 
&\multirow{2}{*}{\textbf{\footnotesize MDS-based}} & \textbf{\footnotesize Statistical} &
\multirow{2}{*}{\textbf{\footnotesize Available}} \\
& & \textbf{\footnotesize Ensemble}
 & & \textbf{\footnotesize model} & \\
\hline
\hline
\scriptsize{\cite{dekker:capturing}} & & C & \CHECK & & \\
\scriptsize{\cite{duan:three-dimensional}} &  & C & \CHECK & & \CHECK \\
\scriptsize{\cite{tanizawa:mapping}} & & C & \CHECK & & \\
\scriptsize{\cite{ay:three-dimensional}} & & C & \CHECK & & \\
\scriptsize{\cite{ben-elazar:spatial}} & & C & \CHECK & & \CHECK \\
\scriptsize{\cite{varoquaux:statistical}} & Pastis & C & & \CHECK & \CHECK\\
\scriptsize{\cite{bau:three-dimensional}} & & E & & &  \\
\scriptsize{\cite{umbarger:three-dimensional}} & & E  & & &\\
\scriptsize{\cite{zhang:inference}} & chromSDE & C &  \CHECK & & \CHECK\\
\scriptsize{\cite{rousseau:three}} & & E & & \CHECK & \CHECK\\
\scriptsize{\cite{hu:bayesian}} & Bach & E/C &  & \CHECK & \CHECK\\
\scriptsize{\cite{kalhor:genome}} & & E &   & &\\
\scriptsize{\cite{lesne:3d}} & ShRec3D & C  & \CHECK & & \CHECK \\
\scriptsize{\cite{trieu:large}} & & C & & & \\
\scriptsize{\cite{trieu:3D}} & & C & &  & \\
\scriptsize{\cite{trieu:MOGEN}} & MOGEN & C & & & \\
\scriptsize{\cite{nagano:single-cell}} & & E & & & \\
\scriptsize{\cite{paulsen:manifold}} & & C & \CHECK & & \CHECK \\
\hline
\end{tabular}
\end{center}
\caption{\bf A comparison of 3D inference methods}{In this table, we summarize
properties of published methods to infer the 3D structure of the genome: (1)
is it a consensus or a ensemble based inference? (2) Is it an MDS based
method? (3) or relies on a statistical modeling; (4) is the software
available or not (to the best of our knowledge).}
\end{table}

\section*{Model-driven approach: modeling DNA as a flexible fibers under constraints}

@@

The budding yeast \textit{S. cerivisiae}'s 3D structure has been extensively
studied, both through 3C-type studies \citep{dekker:capturing,
duan:three-dimensional, burton:species-level} and through bio-imaging experiments
\citep{berger:high}. The small size of its genome, the well-known hallmarks of its
genome architecture and the availability of high resolution contact maps and
FISH data set quickly led several teams to investigate the minimal set of
constraints needed to reproduce the hallmarks of its genome architecture.

\cite{tjong:physical, tokuda:dynamical, tjong:physical} model {\em S.
cerevisiae}'s chromosomes as flexible random fibers under a small set of
constraints. While the exact modeling proposed by the three groups differ, the
set of constraints can roughly be summarized as: \textit{(i)} the chromosomes
are constrained into a spherical ball representing the nucleus; \textit{(ii)}
centromeres are constrained into a spherical ball tethered to the nuclear
membrane; \textit{(iii)} telomeres are tethered to the nuclear membraine;
\textit{(iv)} rDNA is constrained into the nucleolus, represented as a
spherical ball opposite to the centromeres.

@@

@@

@@

In previous section, I have reviewed data-driven methods to infer either
\textit{consensus} or \textit{ensemble} models of the 3D structure. The reader
may well ask why go through the effort to obtain such model and not directly
study the contact maps?  In this section, I will review a number of downstream
analysis one can do on 3D models, highlighting but limiting myself to results
on the 3D structure of {\em P. falciparum}.

\subsection{Probing the 3D structure of the {\em P. falciparum}}
In the past year, two groups have published studies of {\em P.
falciparum}'s 3D genome architecture using Hi-C. \citet{lemieux:genome-wide}
studied several strains associated with populations expression unique
\textit{var} genes, in order to study key folding properties of the 3D
structure relating to \textit{var} gene expression. On the other hand,
\citet{ay:three-dimensional} focus three key time-point of the development
cycle. Table~\ref{tab:data} summarizes the available data sets and their key
properties after processing the data as described in
\citet{ay:three-dimensional} and combining all available libraries for each
life-cycle stage and strain.

\begin{table}
\scriptsize
\centering
\begin{tabular}{cccccccc}
\hline
\multirow{2}{*}{\textbf{Name}} & \multirow{2}{*}{\textbf{Strain}} &
\multirow{2}{*}{\textbf{Stage}} & \multirow{2}{*}{\textbf{Resolution}} &
\textbf{Number} & \textbf{Perc} & \textbf{Perc}& \multirow{2}{*}{\textbf{Reference}} \\
& & & & \textbf{of contacts} & \textbf{of \textit{cis}}& \textbf{of trans}& \\
\hline
\hline
Ay-\textit{rings} & 3D7 & Late Rings & 10~kb & 16711552 & 43\% & 57\% &  \cite{ay:three-dimensional} \\
Ay-\textit{trophozoites} & 3D7 & Trophozoites &10~kb & 56348498 & 53\% & 47\% & \cite{ay:three-dimensional} \\
Ay-\textit{schizonts} & 3D7 & Schizonts & 10~kb & 11652832 & 55\% & 45 \% & \cite{ay:three-dimensional} \\
Lemieux-\textit{A4+} & IT/BC6+ & Rings & 25~kb & 18488252 & 19\% & 81\% & \cite{lemieux:genome-wide} \\
Lemieux-\textit{A4} & IT/3G8 & Rings &  25~kb & 19674672 & 28\% & 72\% & \cite{lemieux:genome-wide}\\
Lemieux-\textit{A44} & IT/BC6- & Rings & 25~kb & 18660594 & 25\% & 75\% & \cite{lemieux:genome-wide}\\
Lemieux-\textit{DCJ\_On} & NF54/DCJ on & Rings & 25~kb & 3098370 & 26\% & 74\% &\cite{lemieux:genome-wide} \\
Lemieux-\textit{DCJ\_Off} & NF54/DCF Off & Rings & 25~kb & 2533470 & 26\% & 73\% &\cite{lemieux:genome-wide} \\
Lemieux-\textit{B15C2} & NF54/B15C2 & Rings & 25~kb &  1022996 & 12\% & 88\% & \cite{lemieux:genome-wide}\\
\hline
\end{tabular}
\caption{Summary of the available {\em P. falciparum} Hi-C datasets}
\end{table}

\fixme{Add the fact we have time points}

\subsection*{Model evaluation}

A substantial difficulty in modelling the 3D structure of the genome is that
model evaluation tends to the subjective. What is the relevant measure?
``Truth'' is generally not fully available, except for a few pairs of loci or
in simulations. Is validating the colocalization of a pair or a few pairs of
loci via FISH experiments enough? Is fit to the contact maps or agreements
between modeling techniques relevant?

\fixme{Add section on model evaluation}

\subsection*{Structure stability across time points, clustering and other
variance analysis \dots}

The reader may well ask how sensitive the result is either initialization or
how. In short, it is not. To study the stability of the results with respect
to the initialization, I perform for each stage presented here 1000
optimization, starting from different initialization.

The first question that arise is simply: are structures from the same time
points but from a different start point more alike than structures from
different time points? To answer this question, I compute for each structure
and each time point a feature matrix, composed of a subsampled pairwise
euclidean distances of each structure (so that the resulting feature matrix
fits in memory on a laptop). I then compute a stochastic PCA on this feature
matrix and plot the projection in 2D space using the 3 first components. The
results clearly show that structures from the same time point cluster
strongly. Repeating this experiment on singly chromosomes yields the same
result.

The second question one can ask is how do the structures differ. Tackling this
question is very challenging, but can be reformulated in a much easier way.
Are the hallmarks of interest conserved across structures of a same stage?
\citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both identified
{\em P. falciparum} folded in very specific ways, with VRSSM genes highly
interacting. \citet{ay:three-dimensional} also observed strong clustering of
the centromeres, and enrichment in interaction at the telomeres. These
observation can lead to a rigorous approach at identifying whether families
of loci clustered in the structures and even quantifying this in a rigourous
manner.

\subsection{Chromatin compaction}

Bau et al

@@

To assess whether groups of genes are colocolized in a 3D model,
\citet{ay:three-dimensional} leveraged a statistical method developed by
\citet{witten:assessment}, which requires labeling each pair of loci in two
groups: "close" or "far".The authors used varying distance thresholds (10\%,
20\% and 40\% of the nuclear diameter) to deem a locus pair “close” and
labeled all remaining pairs in the set as “far”.  The authors then compare the
enrichment of loci pairs of a group being close and far by resampling loci
among a same chromosome.

@@

Applying MEPD to the Ring stage, \citet{capurso:distance-based} confirm that
centromeres, telomeres, VRSM genes (both overall, subtelomeric and internal)
colocalize.

@@

Last but not least, an exciting contribution of \citet{ay:three-dimensional}
is the integrative analysis of the 3D structure with gene expression using an
unsupervised learning technique called ``kernel Canonical Correlation
Analysis'' \citep{bach:kernel}. In this particular case,  KernelCCA aims at
extracting a set of orthogonal gene expression component that exhibit
coherence with respect (here) to the 3D structure. It can be helpful to
think of this procedure as performing a Principal Component Analysis on the
set of gene expression profiles but such that the component extracted are
correlated with the gene expression.

Let us take a closer look at how formally kCCA is used in this context.
Consider the set of $n$ genes $g \in \mathcal{G}$. Each gene $g$ is
represented on the one hand by its gene expression profile $e(g) \in \RR^{p}$
and its 3D position $x(g) \in \RR^3$. Assume the set of gene expression
profiles is mean centered and of unit variance.

The goal of the procedure is to extract a gene expression component $v\in
\RR^p, \|v\|=1$ such that it is both representative of set of gene expression profile,
but also correlated with the 3D structure. We thus construct two scores to
assess those properties. First, define $V(v) = \underset{g \in
\mathcal{G}}{\sum} (v^T e(g))^2$. $V(v)$ measures the percentage of variance
explained among the gene expression profile when projected onto the component
$v$. Note that maximizing this score results in finding the first principal
component. To assess smoothness along the structure,
\citet{ay:three-dimensional} leverage a standard approach in kernel methods,
where the smoothness of a score $f$ is quantified by the function:

\begin{equation}
S(f) = \frac{f^TK^{-1}_{3D}f}{\|f\|}\,,
\end{equation}

where $K_3D$ is the Gaussian kernel matrix of the genes' 3D coordinates. The
smaller $S(f)$, the smoother $f$ is with respect to the 3D structure.
